---
title: "p8105_hw3_zn2220"
author: "Ziang Niu"
date: "2025-10-02"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
library(p8105.datasets)
library(lubridate)
library(janitor)
library(knitr)
library(ggplot2)
library(forcats)
```

# Problem 1

## 1.0 

### To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations.

Introducing data.

```{r}
data("instacart")
instacart <- janitor::clean_names(instacart)
```

Monitor the data. It has `r nrow(instacart)` rows and `r ncol(instacart)` cols. Let's glimpse at it and make a slice. Some key variables and example are shown in the kable. 

```{r}
glimpse(instacart)

set.seed(235)
instacart_examples <- instacart %>% 
  slice_sample(n = 5) %>% 
  select(order_id, product_name, aisle, department, order_dow, order_hour_of_day)

knitr::kable(instacart_examples, caption = "Illustrative sample of Instacart orders")
```

Important variables include order_id (unique identifier for each order), product_name (the specific item purchased), aisle (store aisle), department (broader store department), order_dow (day of week; 0 = Sunday, 6 = Saturday), and order_hour_of_day (hour of day).
Illustrative examples highlight how these variables describe customer behavior:

In order 3258129, the customer purchased Organic Extra Firm Tofu from the tofu meat alternatives aisle in the deli department, placed on Sunday (day 0) at 3 PM.

In order 2741482, Organic Powdered Sugar was bought from the baking ingredients aisle in the pantry department, placed on Wednesday (day 3) at 11 AM.

In order 651769, Special K Protein Milk Chocolate Protein Shakes came from the protein meal replacements aisle in the personal care department, placed on Wednesday (day 3) at 5 PM.

In order 859477, Variety Pack was purchased from the food storage aisle in the household department, placed on Tuesday (day 2) at 9 AM.

In order 1788253, Organic Grape Tomatoes came from the packaged vegetables fruits aisle in the produce department, placed on Sunday (day 0) at 4 PM.

## 1.1 

### How many aisles are there, and which aisles are the most items ordered from?

Count the total number of aisles and identify the one with the most items ordered.

```{r}
aisle_counts <- instacart %>%
  count(aisle, name = "n_items") %>%
  arrange(desc(n_items))

n_aisles <- nrow(aisle_counts)

top_aisle <- aisle_counts %>%
  slice(1)
```

It has `r n_aisles` aisles, the tibble below shows the the one with the most items ordered.

```{r}
top_aisle
```

## 1.2

### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
aisle_counts %>%
  filter(n_items > 10000) %>%
  ggplot(aes(
    x = fct_reorder(aisle, n_items), 
    y = n_items
  )) +
  geom_col(fill = "#1f77b4", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Top Instacart Aisles by Number of Items Ordered (> 10,000)",
    x = "Aisle Name (Sorted by Order Count)",
    y = "Total Number of Items Ordered",
    caption = "Source: Instacart Data"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color = "grey50")
  )
```






