---
title: "p8105_hw3_zn2220"
author: "Ziang Niu"
date: "2025-10-02"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
library(p8105.datasets)
library(lubridate)
library(janitor)
library(knitr)
library(ggplot2)
library(forcats)
library(dplyr)
```

# Problem 1

## 1.0 

### To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations.

Introducing data.

```{r}
data("instacart")
instacart <- janitor::clean_names(instacart)
```

Monitor the data. It has `r nrow(instacart)` rows and `r ncol(instacart)` cols. Let's glimpse at it and make a slice. Some key variables and example are shown in the kable. 

```{r}
glimpse(instacart)

set.seed(235)
instacart_examples <- instacart %>% 
  slice_sample(n = 5) %>% 
  select(order_id, product_name, aisle, department, order_dow, order_hour_of_day)

knitr::kable(instacart_examples, caption = "Illustrative sample of Instacart orders")
```

Important variables include order_id (unique identifier for each order), product_name (the specific item purchased), aisle (store aisle), department (broader store department), order_dow (day of week; 0 = Sunday, 6 = Saturday), and order_hour_of_day (hour of day).
Illustrative examples highlight how these variables describe customer behavior:

In order 3258129, the customer purchased Organic Extra Firm Tofu from the tofu meat alternatives aisle in the deli department, placed on Sunday (day 0) at 3 PM.

In order 2741482, Organic Powdered Sugar was bought from the baking ingredients aisle in the pantry department, placed on Wednesday (day 3) at 11 AM.

In order 651769, Special K Protein Milk Chocolate Protein Shakes came from the protein meal replacements aisle in the personal care department, placed on Wednesday (day 3) at 5 PM.

In order 859477, Variety Pack was purchased from the food storage aisle in the household department, placed on Tuesday (day 2) at 9 AM.

In order 1788253, Organic Grape Tomatoes came from the packaged vegetables fruits aisle in the produce department, placed on Sunday (day 0) at 4 PM.

## 1.1 

### How many aisles are there, and which aisles are the most items ordered from?

Count the total number of aisles and identify the one with the most items ordered.

```{r}
aisle_counts <- instacart %>%
  count(aisle, name = "n_items") %>%
  arrange(desc(n_items))

n_aisles <- nrow(aisle_counts)

top_aisle <- aisle_counts %>%
  slice(1)
```

It has `r n_aisles` aisles, the tibble below shows the the one with the most items ordered.

```{r}
top_aisle
```

## 1.2

### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
aisle_counts %>%
  filter(n_items > 10000) %>%
  ggplot(aes(
    x = fct_reorder(aisle, n_items), 
    y = n_items
  )) +
  geom_col(fill = "#1f77b4", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Top Instacart Aisles by Number of Items Ordered (> 10,000)",
    x = "Aisle Name (Sorted by Order Count)",
    y = "Total Number of Items Ordered",
    caption = "Source: Instacart Data"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color = "grey50")
  )
```

## 1.3

### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
target_aisles <- c("baking ingredients", "dog food care", "packaged vegetables fruits")

popular_items <- instacart %>%
  filter(aisle %in% target_aisles) %>%
  count(aisle, product_name, name = "order_count") %>%
  group_by(aisle) %>%
  slice_max(order_count, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(aisle, desc(order_count))

kable(popular_items, caption = "Top 3 Most Popular Items in Selected Aisles")
```

## 1.4

### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
target_products <- c("Pink Lady Apples", "Coffee Ice Cream")

mean_hour_table <- instacart %>%
  filter(product_name %in% target_products) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_order_hour = mean(order_hour_of_day, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_order_hour
  ) %>%
  # --- FIX IS HERE: Apply round only to numeric columns ---
  mutate(across(where(is.numeric), ~round(., 2)))

day_names <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
# Note: You might need to adjust the column indexing [2:8] if the product_name column is implicitly kept as the first column.
# The `pivot_wider` output will have 'product_name' followed by the 7 'order_dow' columns.
colnames(mean_hour_table)[2:8] <- day_names

kable(mean_hour_table, caption = "Mean Order Hour of Day by Product and Day of Week")
```

# Problem 2

## 2.1

### There are 116 months between January 2015 and August 2024. How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?

```{r}
df_zori <- read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv")

metadata_cols_count <- 9
data_cols <- (metadata_cols_count + 1):ncol(df_zori)
total_months <- length(data_cols)

df_zori$Observation_Count <- rowSums(!is.na(df_zori[, data_cols]))

zip_codes_116 <- sum(df_zori$Observation_Count == total_months)
zip_codes_less_than_10 <- sum(df_zori$Observation_Count < 10)

cat("Total Months in Analysis Period:", total_months, "\n")
cat("----------------------------------\n")
cat("Number of ZIP codes observed", total_months, "times (every month):", zip_codes_116, "\n")
cat("Number of ZIP codes observed fewer than 10 times:", zip_codes_less_than_10, "\n")

df_results <- df_zori %>% 
  select(RegionName, CountyName, Observation_Count) %>%
  arrange(desc(Observation_Count))

print("\nSample of ZIP Codes and their Observation Counts:")
print(head(df_results))
```

ZIP codes observed fewer than 10 times, representing a small, illiquid rental market or an area dominated by non-rental housing (e.g., single-family owned homes). Reasons for rare observations include low rental volume (not enough listings for a valid index), data sparsity/suppression (failing to meet minimum transaction thresholds set to ensure data quality or protect privacy), and housing type (the area primarily consists of single-family homes or other non-rental properties).

## 2.2

### Create a reader-friendly table showing the average rental price in each borough and year (not month). Comment on trends in this table.

```{r}
borough_mapping <- c(
  "Kings County" = "Brooklyn",
  "Queens County" = "Queens",
  "New York County" = "Manhattan",
  "Bronx County" = "Bronx",
  "Richmond County" = "Staten Island"
)

df_long_zori <- df_zori %>%
  select(CountyName, starts_with("20")) %>%
  pivot_longer(
    cols = starts_with("20"),
    names_to = "Date",
    values_to = "ZORI"
  ) %>%
  mutate(
    Date = as.Date(Date, format = "%Y-%m-%d"),
    Year = as.integer(format(Date, "%Y")),
    Borough = str_replace(CountyName, " County", ""),
    Borough = recode(Borough, !!!borough_mapping)
  ) %>%
  filter(!is.na(Borough) & !is.na(ZORI))

average_rent_by_borough_year <- df_long_zori %>%
  group_by(Borough, Year) %>%
  summarise(
    Average_Rental_Price = mean(ZORI, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(Average_Rental_Price = round(Average_Rental_Price, 2)) %>%
  pivot_wider(
    names_from = Year,
    values_from = Average_Rental_Price
  ) %>%
  arrange(desc(`2024`)) 

print("Average Annual Rental Price (ZORI) by Borough (USD):")
print(average_rent_by_borough_year)
```

The rental market analysis for NYC boroughs from 2015 to 2024 reveals stark contrasts and rapid changes: Manhattan consistently commands the highest rent, surpassing $4,000 by 2024, while the Bronx remains the most affordable. During the initial pandemic years (2020–2021), rents in the denser boroughs (Manhattan and Brooklyn) stalled or decreased, but the Bronx saw a slight increase, suggesting a shift toward lower-cost and less-dense areas. Critically, starting in 2022, all boroughs experienced a sharp, rapid rebound in rent growth, indicating a strong post-pandemic resurgence in demand across the entire NYC rental market.

## 2.3

### Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.

