---
title: "p8105_hw3_zn2220"
author: "Ziang Niu"
date: "2025-10-02"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
library(p8105.datasets)
library(lubridate)
library(janitor)
library(knitr)
library(patchwork)
library(RNHANES)
```

# Problem 1

## 1.0 

### To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations.

Introducing data.

```{r}
data("instacart")
instacart <- janitor::clean_names(instacart)
```

Monitor the data. It has `r nrow(instacart)` rows and `r ncol(instacart)` cols. Let's glimpse at it and make a slice. Some key variables and example are shown in the kable. 

```{r}
glimpse(instacart)

set.seed(235)
instacart_examples <- instacart %>% 
  slice_sample(n = 5) %>% 
  select(order_id, product_name, aisle, department, order_dow, order_hour_of_day)

knitr::kable(instacart_examples, caption = "Illustrative sample of Instacart orders")
```

Important variables include order_id (unique identifier for each order), product_name (the specific item purchased), aisle (store aisle), department (broader store department), order_dow (day of week; 0 = Sunday, 6 = Saturday), and order_hour_of_day (hour of day).
Illustrative examples highlight how these variables describe customer behavior:

In order 3258129, the customer purchased Organic Extra Firm Tofu from the tofu meat alternatives aisle in the deli department, placed on Sunday (day 0) at 3 PM.

In order 2741482, Organic Powdered Sugar was bought from the baking ingredients aisle in the pantry department, placed on Wednesday (day 3) at 11 AM.

In order 651769, Special K Protein Milk Chocolate Protein Shakes came from the protein meal replacements aisle in the personal care department, placed on Wednesday (day 3) at 5 PM.

In order 859477, Variety Pack was purchased from the food storage aisle in the household department, placed on Tuesday (day 2) at 9 AM.

In order 1788253, Organic Grape Tomatoes came from the packaged vegetables fruits aisle in the produce department, placed on Sunday (day 0) at 4 PM.

## 1.1 

### How many aisles are there, and which aisles are the most items ordered from?

Count the total number of aisles and identify the one with the most items ordered.

```{r}
aisle_counts <- instacart %>%
  count(aisle, name = "n_items") %>%
  arrange(desc(n_items))

n_aisles <- nrow(aisle_counts)

top_aisle <- aisle_counts %>%
  slice(1)
```

It has `r n_aisles` aisles, the tibble below shows the the one with the most items ordered.

```{r}
top_aisle
```

## 1.2

### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
aisle_counts %>%
  filter(n_items > 10000) %>%
  ggplot(aes(
    x = fct_reorder(aisle, n_items), 
    y = n_items
  )) +
  geom_col(fill = "navy", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Top Instacart Aisles by Number of Items Ordered (> 10,000)",
    x = "Aisle Name (Sorted by Order Count)",
    y = "Total Number of Items Ordered",
    caption = "Source: Instacart Data"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color = "grey50")
  )
```

## 1.3

### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
target_aisles <- c("baking ingredients", "dog food care", "packaged vegetables fruits")

popular_items <- instacart %>%
  filter(aisle %in% target_aisles) %>%
  count(aisle, product_name, name = "order_count") %>%
  group_by(aisle) %>%
  slice_max(order_count, n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(aisle, desc(order_count))

kable(popular_items, caption = "Top 3 Most Popular Items in Selected Aisles")
```

## 1.4

### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
target_products <- c("Pink Lady Apples", "Coffee Ice Cream")

mean_hour_table <- instacart %>%
  filter(product_name %in% target_products) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_order_hour = mean(order_hour_of_day, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_order_hour
  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

day_names <- c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")

colnames(mean_hour_table)[2:8] <- day_names

kable(mean_hour_table, caption = "Mean Order Hour of Day by Product and Day of Week")
```

# Problem 2

## 2.0

Clean and see, Transformation and Combination.

```{r message=FALSE}
zillow_rent_data <- read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv")
zip_code_info <- read_csv("zillow_data/Zip Codes.csv")

glimpse(zillow_rent_data)
summary(zillow_rent_data)
sum(is.na(zillow_rent_data$RegionName))

glimpse(zip_code_info)
sum(duplicated(zip_code_info$ZipCode))
sum(is.na(zip_code_info$Neighborhood))

final_zillow_data <- zillow_rent_data %>%
  pivot_longer(
    cols = starts_with("20"),
    names_to = "period",
    values_to = "value"
  ) %>%
  clean_names() %>%
  mutate(
    county = str_remove(county_name, " County$"),
    postal_code = as.character(region_name),
    period = ymd(period)
  ) %>%
  left_join(
    zip_code_info %>%
      clean_names() %>%
      mutate(postal_code = as.character(zip_code)),
    by = c("postal_code", "county")
  ) %>%
  select(
    postal_code,
    borough = county,
    area = neighborhood,
    period,
    value
  ) %>%
  arrange(postal_code, period)

write_csv(final_zillow_data, "final_zillow_data.csv")
```

## 2.1

### There are 116 months between January 2015 and August 2024. How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?

```{r}
fz_df <- read_csv("final_zillow_data.csv")

counts <- fz_df %>%
  group_by(postal_code) %>%
  summarize(n_months = n_distinct(period), .groups = "drop")

n_116 <- sum(counts$n_months == 116)
n_less_10 <- sum(counts$n_months < 10)

n_116
n_less_10
```

Some ZIP codes are observed in all 116 months because Zillow consistently tracks rental data for those stable, well-established areas. Others appear rarely because data collection began later, ended earlier, or the ZIP codes represent small, newly developed, or discontinued regions. Missing or incomplete data may also result from changes in ZIP boundaries or limited housing market activity.

## 2.2

### Create a reader-friendly table showing the average rental price in each borough and year (not month). Comment on trends in this table.

```{r}
borough_year_table <- fz_df %>%
  mutate(year = year(period)) %>%
  group_by(borough, year) %>%
  summarize(mean_rent = mean(value, na.rm = TRUE), .groups = "drop") %>%
  arrange(borough, year)

kable(
  borough_year_table,
  digits = 0,
  caption = "Average Rental Price by Borough and Year"
)

```

Across boroughs, rental prices generally show a gradual upward trend from 2015 to early 2020, reflecting steady market growth. Many areas exhibit a noticeable dip around 2020–2021, likely due to the COVID-19 pandemic’s economic effects. After 2021, prices rebound sharply, reaching their highest levels by 2023–2024, indicating post-pandemic recovery and renewed demand, especially in central boroughs like Manhattan and Brooklyn.

## 2.3

### Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.

```{r}
ggplot(fz_df, aes(x = period, y = value, group = postal_code, color = borough)) +
  geom_line(alpha = 0.3) +
  labs(
    title = "NYC Rental Prices by ZIP Code (2015–2024)",
    x = "Year",
    y = "Rental Price ($)",
    color = "Borough",
    caption = "Source: Zillow Observed Rent Index (ZORI)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

```

The plot shows distinct borough-level differences in rental prices. Manhattan maintains the highest rent levels throughout the period, followed by Brooklyn and Queens. A sharp decline is visible around 2020–2021 across most ZIP codes, reflecting the pandemic-driven drop in housing demand. Afterward, rents rise steadily again, with a pronounced rebound in 2022–2024. Outer boroughs such as the Bronx and Staten Island display lower but more stable price trajectories compared to central boroughs.

## 2.4

### Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs; put differently, your plot should facilitate the comparison of the distribution of average rental prices across boroughs. Comment on this plot.

```{r}
rent_2023 <- fz_df %>%
  filter(year(period) == 2023) %>%
  group_by(borough, postal_code) %>%
  summarize(avg_rent_2023 = mean(value, na.rm = TRUE), .groups = "drop")

ggplot(rent_2023, aes(x = borough, y = avg_rent_2023, fill = borough)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.3) +
  labs(
    title = "Distribution of ZIP-Code-Level Average Rental Prices by Borough (2023)",
    x = "Borough",
    y = "Average Monthly Rent ($)",
    caption = "Source: Zillow Observed Rent Index (ZORI)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  )
```

The 2023 distribution plot reveals clear rent disparities among boroughs. Manhattan shows the highest median and widest spread of rents, indicating both high overall costs and large variability between ZIP codes. Brooklyn follows with moderately high rents and a narrower spread. Queens displays mid-range prices with moderate variation, while the Bronx and Staten Island have distinctly lower medians and tighter distributions, suggesting more affordable and homogeneous rental markets. Overall, the figure highlights the strong geographic stratification of housing costs within New York City.

## 2.5

### Combine the two previous plots into a single graphic, and export this to a results folder in your repository.

```{r}
fz_df <- read_csv("final_zillow_data.csv")

p1 <- ggplot(fz_df, aes(x = period, y = value, group = postal_code, color = borough)) +
  geom_line(alpha = 0.3) +
  labs(
    title = "NYC Rental Prices by ZIP Code (2015–2024)",
    x = "Year",
    y = "Rental Price ($)",
    color = "Borough"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )

rent_2023 <- fz_df %>%
  filter(year(period) == 2023) %>%
  group_by(borough, postal_code) %>%
  summarize(avg_rent_2023 = mean(value, na.rm = TRUE), .groups = "drop")

p2 <- ggplot(rent_2023, aes(x = borough, y = avg_rent_2023, fill = borough)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.3) +
  labs(
    title = "Distribution of ZIP-Code-Level Average Rental Prices by Borough (2023)",
    x = "Borough",
    y = "Average Monthly Rent ($)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  )

combined_plot <- p1 / p2
ggsave("results/nyc_rent_combined_plot.png", combined_plot, width = 10, height = 10, dpi = 300)

```

# Problem 3

## 3.1

### Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r}
covar <- read_csv("https://p8105.com/data/nhanes_covar.csv", skip = 4) %>%
  clean_names() %>%
  mutate(
    sex = factor(sex,
                 levels = c(1, 2),
                 labels = c("male", "female")),
    education = factor(education,
                       levels = c(1, 2, 3),
                       labels = c("less than high school",
                                  "high school equivalent",
                                  "more than high school"))
  ) %>%
  filter(age >= 21) %>%
  drop_na(sex, age, bmi, education)

accel <- read_csv("https://p8105.com/data/nhanes_accel.csv") %>%
  clean_names()

nhanes <- left_join(covar, accel, by = "seqn")

```

## 3.2

### Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

```{r}
edu_sex_table <- nhanes %>%
  count(education, sex, name = "n") %>%
  pivot_wider(names_from = sex, values_from = n) %>%
  mutate(total = male + female) %>%
  kable(
    caption = "Number of Men and Women in Each Education Category",
    col.names = c("Education Level", "Men", "Women", "Total")
  )

edu_sex_table

age_plot <- nhanes %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 20) +
  facet_wrap(~education, nrow = 1) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Age Distributions of Men and Women by Education Level",
    x = "Age (years)",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

age_plot
```







